# Dynamic UI {#sec-chap10}


```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")

library(glossary)
glossary::glossary_path("../glossary-pb/glossary.yml")

```

## Table of content for chapter 04 {.unnumbered}

::::: {#obj-chap03}
:::: {.my-objectives}
::: {.my-objectives-header}
Chapter section list
:::

::: {.my-objectives-container}

-   ~~Working with RStudio projects (empty)~~ (@sec-chap01-1)
-   R Basics (@sec-chap01-2)
-   Fixing errors and getting help (@sec-chap01-3)
-   ~~Organizing your work and making it reproducible (empty)~~ (@sec-chap01-4)
-   An Extended Illustration: Duncan's Occupational-Prestige Regression (@sec-chap01-5)
    - Getting, recoding and showing data (@sec-chap01-5-1)
    - Explorative data analysis (@sec-chap01-5-2)
    - Regression analysis (@sec-chap01-5-3)
    - Regression diagnostics (@sec-chap01-5-4)
- ~~R Functions for Basic Statistics (empty)~~ (@sec-chap01-6)
- Generic functions and their methods (@sec-chap01-7)



:::
::::
:::::






## Glossary Entries {#unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::

