# Dynamic UI {#sec-chap10}


```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")

library(glossary)
glossary::glossary_path("../glossary-pb/glossary.yml")

```

## Table of content for chapter 10 {.unnumbered}

::::: {#obj-chap03}
:::: {.my-objectives}
::: {.my-objectives-header}
Chapter section list
:::

::: {.my-objectives-container}


:::
::::
:::::

There are three key techniques for creating dynamic user interfaces:

- Using the update family of functions to modify parameters of input controls.
- Using `tabsetPanel()` to conditionally show and hide parts of the user interface.
- Using `uiOutput()` and `renderUI()` to generate selected parts of the user interface with code.


## Updating inputs

We’ll begin with a simple technique that allows you to modify an input after it has been created: the update family of functions. [Every input control]{.mark}, e.g. `textInput()`, [is paired with an **update function**]{.mark}, e.g. `updateTextInput()`, that allows you to modify the control after it has been created.

The key idea to modify an input is to use `observeEvent()` to trigger the corresponding update<whatever> function whenever the input changes. (For the `observeEvent()` function see @sec-03-observers and @XXX15.3.)

::: {.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-updating-inputs-intro}
: Updating the slider whenever the the minimum or maximum input changes
::::::
:::
::::{.my-r-code-container}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 400
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/01-updating-inputs-intro/app.R >}}
```



::::
:::::

::: 

### Simple uses

#### Reset button


The simplest uses of the update functions are to provide small conveniences for the user. For example, maybe you want to make it easy to reset parameters back to their initial value.

::: {.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-updating-with-reset-button}
: Updating controls with reset button
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 400
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/02-updating-with-reset-button/app.R >}}
```
::::
:::::

::: 


#### Button changed functionality

A similar application is to tweak the text of an action button so you know exactly what it’s going to do.


::: {.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-updating-vutton-text}
: Updating button text if it has changed functionality
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 400
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/03-updating-button-text/app.R >}}
```
::::
:::::

::: 

### Hierarchical select boxes {#sec-10-hierarchical-selct-boxes}

A more complicated, but particularly useful, application of the update functions is to allow interactive drill down across multiple categories. I’ll illustrate their usage with some imaginary data for a sales dashboard.

:::::{.my-resource}
:::{.my-resource-header}
:::::: {#lem-10-kaggle}
: Kaggle --- An online community platform for data scientists and machine learning enthusiasts
::::::
:::
::::{.my-resource-container}
This section uses [Sample Sales Data](https://www.kaggle.com/datasets/kyanyoga/sample-sales-data), a well-documented imaginary dataset provided by `r glossary("Kaggle")` for education, training and research. To download the dataset from Kaggle you have to register. But you can also download the Sample Sales Data as `r glossary("CSV", ".csv")` file from the [GitHub repo of Mastering Shiny](https://github.com/hadley/mastering-shiny/raw/refs/heads/main/sales-dashboard/sales_data_sample.csv) or alternatively my own version of the dataset.

***

[Kaggle](https://www.kaggle.com/) is one of the largest hosting platforms used by data scientists and machine learning enthusiasts globally. It allows users to collaborate with other users, find and publish datasets, use GPU integrated notebooks, and compete with other data scientists to solve data science challenges.  

Founded in 2010 by [Anthony Goldbloom](https://en.wikipedia.org/wiki/Anthony_Goldbloom) and [Jeremy Howard](https://en.wikipedia.org/wiki/Jeremy_Howard_(entrepreneur)), Kaggle was acquired by Google in 2017. 
::::
:::::

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-load-sales-dataset}
: Load and display sales dataset
::::::
:::
::::{.my-r-code-container}
```{r}
#| label: load-sales-dataset

sales <- vroom::vroom("data/sales_data_sample.csv", col_types = list(), na = "")
sales  |> 
  dplyr::select(TERRITORY, CUSTOMERNAME, ORDERNUMBER, dplyr::everything()) |> 
  dplyr::arrange(ORDERNUMBER)
```

::::
:::::

For this demo, I’m going to focus on a natural hierarchy in the data:

- Each territory contains customers.
- Each customer has multiple orders.
- Each order contains rows.

I want to create a user interface where you can:

- Select a territory to see all customers.
- Select a customer to see all orders.
- Select an order to see the underlying rows.

The essence of the UI is simple: I’ll create three select boxes and one output table. The choices for the `customername` and `ordernumber` select boxes will be dynamically generated, so I set `choices = NULL`.

In the server function, there is th following top-down procedure to follow:

:::::{.my-procedure}
:::{.my-procedure-header}
:::::: {#prp-10-replicate-hierarchical-data-structure}
: Replicating the hierarchical data structure in the server function
::::::
:::
::::{.my-procedure-container}

1. Create a reactive, `territory()`, that contains the rows from sales that match the selected territory.
2. Whenever `territory()` changes, update the list of choices in the `input$customername` select box.
3. Create another reactive, `customer()`, that contains the rows from `territory()` that match the selected customer.
4. Whenever `customer()` changes, update the list of choices in the `input$ordernumber` select box.
5. Display the selected orders in `output$data`.

::::
:::::

:::{.column-page-inset}

:::::{.my-example}
:::{.my-example-header}
:::::: {#exm-10-updating-nested-user-input}
: Hierarchical select boxes
::::::
:::
::::{.my-example-container}

::: {.panel-tabset}

###### First take

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-updating-nested-user-input1}
: Hierarchical select boxes: Small demo
::::::
:::
::::{.my-r-code-container}
::: {#lst-10-updating-nested-user-input1}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/04-updating-nested-user-input1/app.R >}}
```

Small demo of hierarchical select boxes
:::

::::
:::::


###### Improved demo

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-updating-nested-user-input2}
: Hierarchical select boxes: More fleshed out
::::::
:::
::::{.my-r-code-container}
::: {#lst-10-updating-nested-user-input2}


```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]


## file: app.R
{{< include apps_10/04-updating-nested-user-input2/app.R >}}

```

Hierarchical select boxes: more fleshed out demo
:::

::::
:::::

:::

::::
:::::

:::

::: {.callout-warning #wrn-10-upating-nested}
###### Couldn't load data without user input

To get a version that works with Shiny and shinylive-r at the same time, I had to provide `fileInput()`. This is suboptimal as the user has to find and select the file to get the intended data for playing around.

It was easy to find a solution for Shiny: Just load the data before the ui/server part into an R object and then reference it from the `server()` function. I couldn’t manage the same result for `shinylive-r`. Local references, relative URL and non-HTTPS URL generate error. The only possible way seems to doanload the data via HTTPS. I tried but didn’t succeed. I asked for help via the [Posit Community](https://forum.posit.co/t/shinylive-in-quarto-document-with-reactive-data-source/204237) and are currently waiting for answers.

:::

A workaround was to provide a file upload control. But even to create a workable demo took me hours. It was a bitter experience to learn that I still didn't understand quite well how reactives work. As the debugger `browser()` did not work, it was finally for me quite helpful to install `print()` outputs in **every** reactive function to see what happens. In the end I solved this problem with three important changes:

1. Creating a reactive value with `sales <- reactiveVal()`.
2. Adding an `observeEvent()` function which fulfilled several tasks:
    - Loading the data after `fileInput()` was active.  
    - Assigning the file to the reactive value `sales`.
    - Updating with the data the choice of the territory `selectInput()` function. (Previously I had to set the UI choices for territory to NULL to prevent an error because the data was not available at start up.)
3. The reference to the previous `sales` object had to be changed to the reactive function `sales()`. Additional had I to add `req(input$upload)` to prevent that the reactive territory function is called immediately after start up without available data.

```{block2, label='success-upload-external-file', type='rmdsuccess'}
In the meanwhile I [received an answer](https://forum.posit.co/t/shinylive-in-quarto-document-with-reactive-data-source/204237/3?u=petzi53). I commit a silly error providing the wrong URL for downloading the raw data.

What follows are the original code chunks from "Mastering Shiny" with a tiny change: Instead to get the salaes data locally I uploaded it via an URL. 

```


:::{.column-page-inset}

:::::{.my-example}
:::{.my-example-header}
:::::: {#exm-10-updating-nested-external-file1}
: Hierarchical select boxes
::::::
:::
::::{.my-example-container}

::: {.panel-tabset}

###### First take

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-updating-nested-external-file1}
: Hierarchical select boxes: Small demo
::::::
:::
::::{.my-r-code-container}
::: {#lst-10-updating-nested-external-file1}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/04-updating-nested-external-file1/app.R >}}
```

Small demo of hierarchical select boxes
:::

::::
:::::


###### Improved demo

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-updating-nested2}
: Hierarchical select boxes: More fleshed out
::::::
:::
::::{.my-r-code-container}
::: {#lst-10-updating-nested-external-file2}


```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [editor, viewer]


## file: app.R
{{< include apps_10/04-updating-nested-external-file2/app.R >}}

```

Hierarchical select boxes: more fleshed out demo
:::

::::
:::::

:::

::::
:::::

:::

### Freezing reactive inputs

Sometimes hierarchical selections can briefly create an invalid set of inputs, leading to a flicker of undesirable output.

:::{.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-freezing-reactive-input1}
: Flicker of undesirable output
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/05-freezing-reactive-input1/app.R >}}
```

::::
:::::

:::

You’ll notice that when you switch datasets the summary output will briefly flicker. That’s because `updateSelectInput()` only has an effect after all outputs and observers have run, so there’s temporarily a state where you have dataset B and a variable from dataset A, so that the output contains summary(NULL).

You can resolve this problem by “freezing” the input with `freezeReactiveValue()`. This ensures that any reactives or outputs that use the input won’t be updated until the next full round of invalidation. It’s good practice to **always** use it when you dynamically change an input value.

:::{.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-freezing-reactive-input2}
: Freezing input with `freezeReactiveValaue()`
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/05-freezing-reactive-input2/app.R >}}
```
::::
:::::

:::

### Circular references

Using an update function to modify value is no different to the user modifying the value by clicking or typing. That means an update function can trigger reactive updates in exactly the same way that a human can. This means that you are now stepping outside of the bounds of pure reactive programming, and you need to start worrying about circular references and infinite loops.


````markdown

library(shiny)

ui <- fluidPage(
    actionButton("action", "start"),
    numericInput("n", "n", 0)
)
server <- function(input, output, session) {
    observeEvent(input$n,
                 updateNumericInput(inputId = "n", value = input$n + 1)
    )
}

shinyApp(ui, server)

````
For example, take the simple app above. It contains a single input control and an observer that increments its value by one. Every time `updateNumericInput()` runs, it changes `input$n`, causing `updateNumericInput()` to run again, so the app gets stuck in an infinite loop constantly increasing the value of `input$n`.

To prevent that the circular reference runs all the time I had to add a start and a stop button. This was a welcome training occasion for me. I used as help the article [Using Action Buttons](https://shiny.posit.co/r/articles/build/action-buttons/).


:::{.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-circular-references}
: Demonstration of circular references
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/06-circular-references2/app.R >}}
```
::::
:::::

:::

### Inter-related inputs

One place where it’s easy to end up with circular references is when you have multiple “sources of truth” in an app. For example, imagine that you want to create a temperature conversion app where you can either enter the temperature in Celsius or in Fahrenheit:

:::{.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-inter-realted-inputs}
: Demo of inter-related inputs
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/07-inter-related-inputs/app.R >}}
```
::::
:::::

:::

If you play around with this app you’ll notice that it mostly works, but you might notice that it’ll sometimes trigger multiple changes. For example:

- Set 120 F, then click the down arrow.
- F changes to 119, and C is updated to 48.
- 48 C converts to 118 F, so F changes again to 118.
- Fortunately 118 F is still 48 C, so the updates stop there.

There’s no way around this problem because you have one idea (the temperature) with two expressions in the app (Celsius and Fahrenheit). Here we are lucky that cycle quickly converges to a value that satisfies both constraints. In general, you are better off avoiding these situations, unless you are willing to very carefully analyse the convergence properties of the underlying dynamic system that you’ve created.

### Exercises

#### Update date


**Task description**: Complete the user interface below with a server function that updates `input$date` so that you can only select dates in `input$year`.


::: {.column-page-inset}

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-10-ex-01-updating-date1}
: Update date input to limit the choices for the selected year
::::::
:::
::::{.my-exercise-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/ex-01-updating-date1/app.R >}}
```


::::
:::::

::: 



#### Update county

**Task description**: Complete the user interface below with a server function that updates `input$county` choices based on `input$state`. For an added challenge, also change the label from “County” to “Parish” for Louisiana and “Borough” for Alaska.


::: {.column-page-inset}

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-10-ex-02-updating-county}
: Update county by chose state
::::::
:::
::::{.my-exercise-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/ex-02-updating-county/app.R >}}
```


::::
:::::

::: 


#### Update country


**Task description**: Complete the user interface below with a server function that updates `input$country` choices based on the `input$continent`. Use `output$data` to display all matching rows.


::: {.column-page-inset}


:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-10-ex-03-updating-country}
: Updating country by chosen state
::::::
:::
::::{.my-exercise-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/ex-03-updating-country/app.R >}}
```

::::
:::::

::: 

#### Updating all


**Task description**: Extend the previous app so that you can also choose to select all continents, and hence see all countries. You’ll need to add `"(All)"` to the list of choices, and then handle that specially when filtering.


::: {.column-page-inset}


:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-10-ex04-updating-country-all}
: Show all countries when continent "All" was chosen
::::::
:::
::::{.my-exercise-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/ex-04-updating-country-all/app.R >}}
```


::::
:::::

::: 

#### Circular reference


**Task description**: What is at the heart of the problem described at <https://community.rstudio.com/t/29307>?


It is a circular reference because the different `numericInput()` are mutually dependent to each other. The first `observeEvent()` listens for `input$A` and changes `input$B`. But the second `observeEvent()` which listens for `input$B` changes `input$A` and therefore triggers the first `observeEvent()` creating an endless circular reference.

## Dynamic visibility

::: {.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-dynamic-visibility-basics}
: Dynamic visibility basics
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 600
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/08-dynamic-visibility-basics/app.R >}}
```


::::
:::::

:::


There are two main ideas here:

- Use tabset panel with hidden tabs.
- Use `updateTabsetPanel()` to switch tabs from the server.

### Conditional UI


::: {.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-conditional-ui}
: Conditional UI
::::::
:::
::::{.my-r-code-container}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/09-conditional-ui/app.R >}}
```

::::
:::::

:::


Note that the value of (e.g.) `input$mean` is independent of whether or not its visible to the user. The underlying HTML control still exists; you just can’t see it.

### Wizard interface

You can also use this idea to create a “wizard”, a type of interface that makes it easier to collect a bunch of information by spreading it across multiple pages. Here we embed action buttons within each “page”, making it easy to go forward and back.

:::{.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-ID-text}
: Numbered R Code Title
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/10-wizard-interface/app.R >}}
```
::::
:::::

:::

Note the use of the `switch_page()` function to reduce the amount of duplication in the server code.

### Exercises

#### Show additonal controls


**Task description**: Use a hidden tabset to show additional controls only if the user checks an “advanced” check box.


:::{.column-page-inset}

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-10-ex-06-show-additional-controls}
: Show additional controls after selecting a check box
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/ex-06-show-additional-controls/app.R >}}
```
::::
:::::

:::


#### Choose geom

Create an app that plots `ggplot(diamonds, aes(carat))` but allows the user to choose which geom to use: `geom_histogram()`, `geom_freqpoly()`, or `geom_density()`. Use a hidden tabset to allow the user to select different arguments depending on the geom: `geom_histogram()` and `geom_freqpoly()` have a `binwidth` argument; `geom_density()` has a `bw` argument.

:::{.column-page-inset}

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-10-choose-geom}
: Choose geom for plotting
::::::
:::
::::{.my-example-container}

::: {.panel-tabset}

###### shiny

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-choose-geom-shiny}
: Choose geom for plotting ({**shiny**} variant)
::::::
:::
::::{.my-r-code-container}
::: {#lst-10-choose-geom-shiny}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/ex-07-choose-geom-shiny/app.R >}}
```

Choose geom for plotting diamonds data. User interface with {**shiny**} 

:::

::::
:::::


###### bslib

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-choose-geom-bslib}
: Choose geom for plotting ({**shiny**} variant)
::::::
:::
::::{.my-r-code-container}
::: {#lst-10-choose-geom-bslib}  
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/ex-07-choose-geom-bslib/app.R >}}
```

Choose geom for plotting diamonds data. User interface with {**bslib**} 
:::

::::
:::::

:::

::::
:::::

:::

::: {.callout-note #nte-10-choose-geom}
###### Limiting the user input to appropriate number ranges or choices

- To get an adequate number range for `geom_histogram()` and `geom_freqpoly()` I had to experiment with the data. 
- For the `bw` argument of the density distribution I had to read (and understand!) the [docs about kernel density estimation](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/bandwidth.html). 

Bandwidth refers to a parameter that controls the smoothness of the estimated density function. It determines the width of the `r glossary("kernel")` used in the estimation process. Specifically, the bandwidth is the standard deviation of the kernel, which influences how much each data point contributes to the overall density estimate. ([What does bandwidth mean?](https://stats.stackexchange.com/questions/61374/what-does-bandwidth-mean) and [Wikipedia about bandwidth selection](https://en.wikipedia.org/wiki/Kernel_density_estimation#Bandwidth_selection))

To experiment for choosing an appropriate bandwidth is important because different bandwidths may provide you with very diverging impressions of the underlying distribution. For more details, read: [The importance of kernel density estimation bandwidth](https://aakinshin.net/posts/kde-bw/).

:::



::: {.callout-caution #cau-10-input-number-decimal-separator}
###### Problem with loalization of decimal separator

Surprisingly I got commas as decimal separator which I couldn't change to dots. I tried it with different approaches:

- Setting the output of decimal separator to a dot using the R command `options(OutDec=".")`. Does not help because it is already the default. ([Option settings](https://stat.ethz.ch/R-manual/R-devel/library/base/html/options.html))
- Setting a different locale for specific Shiny apps with `Sys.setlocale("LC_ALL","C")` or `Sys.setlocale("LC_ALL","en_US.UTF-8")` ([Setting a different locale for specific Shiny apps](https://stackoverflow.com/questions/61656119/setting-a-different-locale-for-specific-shiny-apps-only-on-shinyapps-io)).
- Using a `textInput()` and convert it to a number in R with `as.numeric(input$mynumber)`. You could also change the comma separator to a dot with `as.numeric(sub(",", ".", input$mynumber))` But the problem with `textInput()` is that the user could could type in any string, not just a number. ([Shiny apps_issue with decimal point/comma](https://groups.google.com/g/shinyapps-users/c/HGDNwV5y314)) --- So this approach would require a more complex workaround using input validation, maybe with the {**shinyFeedback**} package.

Instead of `numericInput()` I used as a workaround `sliderInput()`

> The HTML5 input type=number is inadequate from the localization point of view, due to both the definition and the implementations. It is meant to be localized but as per the locale of the browser, which you cannot set or even know as a designer/author. ([Localization of input type number](https://stackoverflow.com/questions/13412204/localization-of-input-type-number))
:::


#### Choose geom advanced {#sec-10-choose-geom-advanced}



## Creating UI with code {#sec-10-creating-ui-with-code}

The update functions only allow you to change existing inputs, and a tabset only works if you have a fixed and known set of possible combinations. Sometimes you need to create different types or numbers of inputs (or outputs), depending on other inputs.

There are two parts to this solution:

- `uiOutput()` inserts a placeholder in your ui. This leaves a “hole” that your server code can later fill in.
- `renderUI()` is called within `server()` to fill in the placeholder with dynamically generated UI.

### Getting started

:::{.column-page-inset}

:::::{.my-example}
:::{.my-example-header}
:::::: {#exm-10-creating-ui-basics}
: Creating UI with code
::::::
:::
::::{.my-example-container}

::: {.panel-tabset}

###### Basic

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-creating-ui-basics1}
: Creating UI with code: basics
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 450
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/11-creating-ui-basics1/app.R >}}
```

::::
:::::


###### Improved

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-creating-ui-bascis2}
: Creating UI with code with values transferred 
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 450
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/11-creating-ui-basics2/app.R >}}
```




::::
:::::

:::

::::
:::::

:::

The use of `isolate()` in the improved version is important. We’ll come back to what it does in @XXX_15.4.1, but here it ensures that we don’t create a reactive dependency that would cause this code to re-run every time `input$dynamic` changes (which will happen whenever the user modifies the value). We only want it to change when `input$type` or `input$label` changes.


::: {.callout-warning #wrn-10-creating-ui-basics2}
The book example does not work. It raises the error:

> In sliderInput(): `min`, `max`, and `value` cannot be NULL, NA, or empty.

The problem is that at app initialization, `input$dynamic` is `NULL` which rise the error. It could be fixed with `value=0` (like default) when `input$dynamic` is `NULL`.

Before I looked at the GitHub Repo of "Mastering Shiny" I tried to solve the problem with reserving slider and numeric input to get the latter as the first (default) option. A numeric input does not raise an error. But this hack does not work if the user changes immediately into slider mode, without to input a numeric value. 

This problem is already reported on the [repo website of Mastering Shiny](https://github.com/hadley/mastering-shiny/issues/515). The author of issues 515 proposed also a solution with the `% || %` [null coalescing operator](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Control.html) I have never heard about. `x % || % y` is a 1-line function and an alternative way to call `if (is.null(x)) y else x`.

(Hadley uses this trick too to solve the same problem in the next section.)
:::


There are two downsides creating UI with code:

- Relying on it too much can create a laggy UI. For good performance, strive to keep fixed as much of the user interface as possible.
- When you change controls, you lose the currently selected value. Maintaining existing state is one of the big challenges of creating UI with code. This is one reason that selectively showing and hiding UI is a better approach if it works for you — because you’re not destroying and recreating the controls, you don’t need to do anything to preserve the values.

However, in many cases, we can fix the problem by setting the value of the new input to the current value of the existing control.



### Multiple controls {#sec-multiple-controls}

Dynamic UI is most useful when you are generating an arbitrary number or type of controls. That means that you’ll be generating UI with code, and I recommend using functional programming for this sort of task. 

Imagine that you’d like users to be able to supply their own color palette. They’ll first specify how many colors they want, and then supply a value for each color. The ui is pretty simple: we have

- a `numericInput()` that controls the number of inputs, 
- a `uiOutput()` where the generated text boxes will go, and 
- a `textOutput()` that demonstrates that we’ve plumbed everything together correctly.

:::{.column-page-inset}

:::::{.my-example}
:::{.my-example-header}
:::::: {#exm-10-creating-ui-multiple-controls}
: Creating UI with multiple controls
::::::
:::
::::{.my-example-container}

::: {.panel-tabset}

###### Basic

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-creating-UI-with-multiple-controls}
: Creating UI with multiple controls
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/12-creating-ui-multiple-controls1/app.R >}}
```

::::
:::::


###### Improved

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-creating-ui-multiple-controls2}
: Numbered R Code Title
::::::
:::
::::{.my-r-code-container}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/12-creating-ui-multiple-controls2/app.R >}}
```
::::
:::::

:::

::::
:::::

:::

Hadley uses here `purrr::map()` and `purrr::reduce()`, but you could certainly do the same with the base `lapply()` and `Reduce()` functions. If you’re not familiar with the `map()` and `reduce()` of functional programming, you might want to take a brief detour to read [Functional programming](https://adv-r.hadley.nz/functionals.html) before continuing. We’ll also come back to this idea in @XXX_18.

Although the server function is short it contains some big ideas:

- It uses a reactive, `col_names()`, to store the names of each of the colour inputs the user is about to generate.
- Using `purrr::map()` to create a list of `textInput()`s, one each for each name in `col_names()`. `renderUI()` then takes this list of HTML components and adds it to UI.
- Using a new trick to access the values  of the input values. So far we’ve always accessed the components of input with `$`, e.g. `input$col1`. But here we have the input names in a character vector, like `var <- "col1"`. `$` no longer works in this scenario, so we need to swich to `[[`, i.e. `input[[var]]`.
- Using `purrr::map_chr()` to collect all values into a character vector, and display that in `output$palette`. Unfortunately there’s a brief period, just before the new inputs are rendered by the browser, where their values are NULL. This causes `purrr::map_chr()` to error, which we fix by using the handy `%||%` function: it returns the right-hand side whenever the left-hand side is NULL.

If you run this app, you’ll discover a really annoying behavior: whenever you change the number of colors, all the data you’ve entered disappears. We can fix this problem by using the same technique as before: setting value to the (isolated) current value. We’ll also tweak the appearance in the improved version to look a little nicer, including displaying the selected colors in a plot.


### Dynamic filtering

The last example in this chapter is to create an app that lets you dynamically filter any data frame. Each numeric variable will get a range slider and each factor variable will get a multi-select, so (e.g.) if a data frame has three numeric variables and two factors, the app will have three sliders.


:::{.column-page-inset}

:::::{.my-example}
:::{.my-example-header}
:::::: {#exm-10-creating-ui-dynamic-filter}
: Creating UI for dynamic filtering of data
::::::
:::
::::{.my-example-container}

::: {.panel-tabset}

###### Basic

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-creating-ui-dynamic-filter1}
: Creating UI for dynamic filtering of data: First take
::::::
:::
::::{.my-r-code-container}

::: {#lst-10-creating-ui-dynamic-filter1}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/13-creating-ui-dynamic-filter1/app.R >}}
```

Creating UI for dynamic filtering of data: First take
:::
::::
:::::


###### Improved

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-creating-ui-dynamic-filter2}
: Creating UI for dynamic filtering of data: Second version
::::::
:::
::::{.my-r-code-container}

::: {#lst-10-creating-ui-dynamic-filter2}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/13-creating-ui-dynamic-filter2/app.R >}}
```

Creating UI for dynamic filtering of data: Improved
:::
::::
:::::

###### Generalized

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-10-creating-ui-dynamic-filter3}
: Creating UI for dynamic filtering of data: Third version
::::::
:::
::::{.my-r-code-container}
::: {#lst-10-creating-ui-dynamic-filter3}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 700
#| components: [editor, viewer]

## file: app.R
{{< include apps_10/13-creating-ui-dynamic-filter3/app.R >}}
```

Creating UI for dynamic filtering of data: Generalized
:::
::::
:::::

:::

::::
:::::

:::

**Basic version**

1. **`make_ui()`**: Creating this app we start with a function `make_ui()` that creates the UI for a single variable. It’ll return a range slider for numeric inputs, a multi-select for factor inputs, and `NULL` (nothing) for all other types.
2. **`filter_var()`**: The next step is to write the server side equivalent of `make-ui()`: `filter_var()` takes the variable and value of the input control, and returns a logical vector saying whether or not to include each observation. Using a logical vector makes it easy to combine the results from multiple columns.

**Improved version**

You might notice that the app only works with three columns. We can make it work with all the columns by using a little functional programming:

- In `ui()` use `purrr::map()` to generate one control for each variable.
- In `server()`, use `purrr::map()` to generate the selection vector for each variable. 
- Then use `reduce()` to take the logical vector for each variable and combine them into a single logical vector by `&`-ing each vector together.

Again, don’t worry too much if you don’t understand exactly what’s happening here. The main take away is that once you master functional programming, you can write very succinct code that generate complex, dynamic apps.

**Generalized version**

From there, it’s a simple generalization to work with any data frame. @lst-10-creating-ui-dynamic-filter3 illustrates it using the data frames in the {**datasets**} package, but you can easily imagine how you might extend this to user uploaded data.


### Dialog boxes

Before we finish up, there is another but related technique to mention: dialog boxes. 

You’ve seen them already in @sec-08-explicit-confirmation, where the contents of the dialog was a fixed text string. But because `modalDialog()` is called from within the server function, you can actually dynamically generate content in the same way as `renderUI()`. This is a useful technique to have in your back pocket if you want to force the user to make some decision before continuing on with the regular app flow.

### Exercises {#sec-10-exercises3}

## Glossary Entries {#unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::

