# Reactive building blocks {#sec-chap15}


```{r}
#| label: setup
#| results: hold
#| include: false

base::source(file = "R/helper.R")

library(glossary)
glossary::glossary_path("../glossary-pb/glossary.yml")

library(shiny)
reactiveConsole(TRUE)
```

## Table of content for chapter 15 {.unnumbered}

::::: {#obj-chap03}
:::: {.my-objectives}
::: {.my-objectives-header}
Chapter section list
:::

::: {.my-objectives-container}


:::
::::
:::::



## Reactive values

There are two types of reactive values:

- A single reactive value, created by `reactiveVal()`.
- A list of reactive values, created by `reactiveValues()`.

They have slightly different interfaces for getting and setting values.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-reactive-values-interface}
: Interfaces for getting and setting reactive values
::::::
:::
::::{.my-r-code-container}

::: {#lst-15-reactive-values-interface}

```{r}
#| label: reactive-values-interface
#| results: hold

x <- reactiveVal(10)
x()       # get
x(20)     # set
x()       # get

r <- reactiveValues(x = 10)
r$x       # get
r$x <- 20 # set
r$x       # get
```

`reactiveVal()` and `reactiveValues()` have different interfaces
:::
::::
:::::

While they look different, they behave the same, so you can choose between them based on which syntax you prefer. In this book Hadley use `reactiveValues()` because the syntax is easier to understand at a glance, but in his own code h tends to use `reactiveVal()` because the syntax makes it’s clear that something weird is going on.

It’s important to note that both types of reactive values have so called reference semantics. Most R objects have copy-on-modify semantics which means that if you assign the same value to two names, the connection is broken as soon as you modify one.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-copy-on-modify-semantics}
: Copy on modify semantics in R
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-copy-on-modify-semantics}
```{r}
#| label: copy-on-modify-semantics

a1 <- a2 <- 10
a2 <- 20
a1 # unchanged
```

Copy on modify semantics in R
:::

::::
:::::

Reactive values do not have copy-on-modify semantics. They always keep a reference back to the same value so that modifying any copy modifies all values:

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-reference-semantics}
: Reference semantics of reactive values
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-reference-semantics}
```{r}
#| label: reference-semantics
#| results: hold

b1 <- b2 <- reactiveValues(x = 10)
glue::glue("Start: b1 and b2 have the same value. b1$x = ", {b1$x}, ", b2$x = ", {b2$x})
b1$x <- 20
glue::glue("Modification: Only b1$x has changed to ", {b1$x}, ", but b2$x has also changed to", {b2$x})

```

Reactive values have reference semantics
:::

::::
:::::

We’ll come back to why you might create your own reactive values in @sec-chap16. Otherwise, most of the reactive values you’ll encounter will come from the input argument to the server function. These are a little different to the `reactiveValues()` that you create yourself because they’re read-only: you can’t modify the values because Shiny automatically updates them based on user actions in the browser.


### Exercises {#sec-15-exercises-1}

#### Two list of reactive values

What are the differences between these two lists of reactive values? Compare the syntax for getting and setting individual reactive values.



:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-15-two-lists-of-reactive-values}
: Exercise: Two lists of reactive values
::::::
:::
::::{.my-exercise-container}

::: {#lst-15-two-lists-of-reactive-values}

```{r}
#| label: two-lists-of-reactive-values
#| results: hold 

l1 <- reactiveValues(a = 1, b = 2)
l2 <- list(a = reactiveVal(1), b = reactiveVal(2))

glue::glue("l1$a = {l1$a}")
glue::glue("l1$b = {l1$b}")
glue::glue("l2$a() = {l2$a()}")
glue::glue("l2$b() = {l2$b()}")

```

Exercise: Two lists of reactive values

:::
***

Content of `l1` are values, content of `l2` are functions
::::
:::::

#### Has `reactiveVal()` reference semantics?

Design and perform a small experiment to verify that `reactiveVal()` also has reference semantics.

:::::{.my-exercise}
:::{.my-exercise-header}
:::::: {#exr-15-reactive-val-reference-semantics}
: Exercise: Has `reactiveVal()` reference semantics?
::::::
:::
::::{.my-exercise-container}

::: {#lst-15-reactiveVal-reference-semantics}


```{r}
#| label: reactive-val-reference-semantics

r1 <- r2 <- reactiveVal(10)
r1(20)
r2()
```
Exercise: Has `reactiveVal()` reference semantics?
:::

***

Yes, `reactiveVal()` also has reference semantics: I initialized r1 & r2 with `reactiveVal(10)`, then I changed only to value of `r1(20)`- Printing the content of `r2()` results also in 20, although I hadn't change `r2()`directly.

::::
:::::

## Reactive expressions

Recall that a reactive has two important properties: it’s lazy and cached. This means that it only does work when it’s actually needed, and if called twice in a row, it returns the previous value.

There are two important details that we haven’t yet covered: what reactive expressions do with errors, and why `base::on.exit()` works inside of them.

### Errors

Reactive expressions cache errors in exactly the same way that they cache values. For example, take this reactive:

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-cache-errors}
: Reactive expressions cache errors
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-cache-errors}
```{r}
#| label: cache-errors1
#| error: true

r <- reactive(stop("Error occured at ", Sys.time(), call. = FALSE))
r()

```

```{r}
#| label: cache-errors2
#| error: true

Sys.sleep(2)
r()
```

Reactive expressions cache errors
:::

::::
:::::

Errors are also treated the same way as values when it comes to the reactive graph: errors propagate through the reactive graph exactly the same way as regular values. The only difference is what happens when an error hits an output or observer:

- An error in an output will be displayed in the app.
- An error in an observer will cause the current session to terminate. If you don’t want this to happen, you’ll need to wrap the code in `try()` or `tryCatch()`.

This same system powers `req()` (@sec-08-cancelling-execution), which emits a special type of error. This special error causes observers and outputs to stop what they’re doing but not otherwise fail. By default, it will cause outputs to reset to their initial blank state, but if you use `req(..., cancelOutput = TRUE)` they’ll preserve their current display.

### `on.exit()`

You can think of `reactive(x())` as a shortcut for `function() x()`, automatically adding laziness and caching. This is mostly of importance if you want to understand how Shiny is implemented, but means that you can use functions that only work inside functions. The most useful of these is `on.exit()` which allows you to run code when a reactive expression finishes, regardless of whether the reactive successfully returns an error or fails with an error. This is what makes `on.exit()` work in @sec-08-removing-on-completion.

### Exercises {#sec-15-exercises-2}

#### Error propagation with `stop()`

Use the {**reactlog**} package to observe an error propagating through the reactives in the following app, confirming that it follows the same rules as value propagation.



:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-error-propagation1}
: Error propagation
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-error-propagation1}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 450
#| components: [editor, viewer]

## file: app.R
{{< include apps_15/01-ex-error-propagation/app.R >}}
```

Analyzing the error propagation using `stop()` with {**reactlog**}
:::

::::
:::::


I added the line `reactlog::reactlog_enable()` to the app code, started the app and ticked the check box "error?". In the console I received 

> Warning: Error in \<reactive:a\>: Error

I started {**reactlog**} with the {{< kbd Cmd-F3 >}} shortcut and analyzed the propagation. I could confirm that the error followed the same rules as the value.

#### Error propagation with `req()`

Modify the above app to use `req()` instead of `stop()`. Verify that events still propagate the same way. What happens when you use the `cancelOutput` argument?


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-error-propagation2}
: Error propagation
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-error-propagation2}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 450
#| components: [editor, viewer]

## file: app.R
{{< include apps_15/02-ex-error-propagation/app.R >}}
```

Analyzing the error propagation using `req()` with {**reactlog**}
:::

::::
:::::

- Using `req()` in the `if()` condition does not `3` as in @lst-15-error-propagation2 out doesn't output anything. As the condition `req() = TRUE` is not fulfilled Shiny stops the execution.
- Using `req(…, cancelOutput = TRUE)` shows for me no difference. It is said the in the documentation that 

> it is treated slightly differently if one or more outputs are currently being evaluated. In those cases, the reactive chain does not proceed or update, but the output(s) are left is whatever state they happen to be in (whatever was their last valid state).

But in the `reactlog` it shows me the new condition (`input$error 'logo TRUE`). But reading the next pargraph in the docs gives an explanation:

> Note that this is always going to be the case if this is used inside an output context (e.g. `output$txt <- ...`). **It may or may not be the case if it is used inside a non-output context** (e.g. `reactive()`, `observe()` or `observeEvent()`) — depending on whether or not there is an `output$...` that is triggered as a result of those calls. (emphasis is mine)

## Observers and Outputs {#sec-15-observers-and-outputs}

Observers and outputs are terminal nodes in the reactive graph. They differ from reactive expressions in two important ways:

- They are eager and forgetful — they run as soon as they possibly can and they don’t remember their previous action. This eagerness is “infectious” because if they use a reactive expression, that reactive expression will also be evaluated.
- The value returned by an observer is ignored because they are designed to work with functions called for their side-effects, like `cat()` or `write.csv()`.

Observers and outputs are powered by the same underlying tool: `observe()`. This sets up a block of code that is run every time one of the reactive values or expressions it uses is updated. [Note that the observer runs immediately when you create it]{.mark} — it must do this in order to determine its reactive dependencies.

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-observe-test}
: Testing behavior of `observe()`
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-observe-test}
```{{r}}
#| label: observe-test
#| results: hold

y <- reactiveVal(10)
observe({
  message("`y` is ", y())
})

y(5)
y(4)
```

Nut run: Testing behavior of `observe()`
:::

::::
:::::

Hadley rarely uses `observe()` in this book, because it’s the low-level tool that powers the user-friendly `observeEvent()`. Generally, you should stick with `observeEvent()` unless it’s impossible to get it to do what you want. In this book, there is only one case where `observe()` is necessary. See @sec-16-pausing-animations

`observe()` also powers reactive outputs. Reactive outputs are a special type of observers that have two important properties:

- They are defined when you assign them into output, i.e. `output$text <- ...` creates the observer.
- They have some limited ability to detect when they’re not visible (i.e. they’re in non-active tab) so they don’t have to recompute.

It’s important to note that `observe()` and the reactive outputs don’t “do” something, but “create” something (which then takes action as needed). That mindset helps you to understand what’s going on in this example:

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-observe-calls-observe}
: `observe()` calls `observe()`
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-observe-calls-observe}
```{r}
#| label: observe-calls-observe
#| results: hold

x <- reactiveVal(1)
observe({
  x()
  observe(print(x()))
})
x(2)
x(3)
```

`observe()` calls `observe()`
:::

::::
:::::

::: {.callout-caution #cau-15-observe-without-value}
In contrast to the book example my observe() functions does not assign a value to `y`. This generates a 
> Warning: Error in y: could not find function "y"

Observers do not yield a result, they are only useful for their side effects.
:::


Each change to `x` causes the observer to be triggered. The observer itself calls `observe()` setting up another observer. So each time `x` changes, it gets another observer, so its value is printed another time.

As a general rule, you should only ever create observers or outputs at the top-level of your server function. If you find yourself trying to nest them or create an observer inside an output, sit down and sketch out the reactive graph that you’re trying to create — there’s almost certainly a better approach. It can be harder to spot this mistake directly in a more complex app, but you can always use the `reactlog`: just look for unexpected churn in observers (or outputs), then track back to what is creating them.

## Isolating code

To finish off the chapter we need to discuss two important tools for controlling exactly how and when the reactive graph is invalidated. In this section, we’ll discuss `isolate()`, the tool that powers `observeEvent()` and `eventReactive()`, and that lets you avoid creating reactive dependencies when not needed. In the next section, you’ll learn about `invalidateLater()` which allows you to generate reactive invalidations on a schedule.

### `isolate()` {#sec-15-isolate}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-infinite-loop-observer}
: Observer with infinite loop
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-infinite-loop-observer}


```{r}
#| label: infinite-loop-observer
#| eval: false

r <- reactiveValues(count = 0, x = 1)
observe({
  r$x
  r$count <- r$count + 1
  on.exit()
  # if (r$count > 10) stop("To many calls") # my addition
  print(r$count)                          # my addition
})
```

Not run: Observer with infinite loop

:::

::::
:::::

If you were to run @lst-15-infinite-loop-observer, you’d immediately get stuck in an infinite loop because the observer will take a reactive dependency on `x` **and** `count`; and since the observer modifies count, it will immediately re-run.

I tried to run @lst-15-infinite-loop-observer with two addition: The first one tries to stop the loop after 10 rounds, the second one prints the result for each run.

But this didn't work. To stop a running process in Shiny is a complex enterprise as I have found out. Searching on the Internet I learned about some pointers to solve this problem. But here I have skipped this non-trivial learning task and prevented to run the infinite loop with `eval = false`

:::::{.my-resource}
:::{.my-resource-header}
:::::: {#lem-15-stop-ruuning-shiny-processes}
: How to stop running Shiny processes without killing the app 
::::::
:::
::::{.my-resource-container}


- [StackOverflow answer](https://stackoverflow.com/a/76005099/7322615) with some important references
- [ExtendendTask](https://rstudio.github.io/shiny/reference/ExtendedTask.html) is a relatively new Shiny function about running long tasks in the background, which could possible also address the stop problem of Shiny process.

Additionally I found reference to other packages which could use to stop running processes:

- [{**callr**}](https://callr.r-lib.org/): Call R from R
- [{**future**}](https://future.futureverse.org/): Unified Parallel and Distributed Processing in R for Everyone
- [{**ipc**}](https://github.com/fellstat/ipc): Tools for Message Passing Between Processes
- [{**mirai**}](https://mirai.r-lib.org/): Minimalist Async Evaluation Framework for R 
::::
:::::

Shiny provides `isolate()` to resolve the problem of undesired reactive dependency. This function allows you to access the current value of a reactive value or expression without taking a dependency on it:

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-using-isolate}
: Accessing reactive value without taking a dependency on it
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-using-isolate}
```{r}
#| label: using-isolate

r <- reactiveValues(count = 0, x = 1)
r$count 

class(r)
observe({
  r$x
  r$count <- isolate(r$count) + 1
})

r$x <- 1
r$x <- 2
r$count

r$x <- 3
r$count
```

Using `isolate()`: Accessing reactive values or reactive expressions without taking a dependency on it
:::

::::
:::::

Like `observe()`, a lot of the time you don’t need to use `isolate()` directly because there are two useful functions that wrap up the most common usage: `observeEvent()` and `eventReactive()`.


### `observeEvent()` and `eventReactive()`

When you saw the code above, you might have remembered @sec-03-observers and wondered why we didn’t use `observeEvent()`:

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-replace-observe-with-observeEvent}
: Replace `observe()` with `observeEvent()`
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-replace-observe-with-observeEvent}

```{r}
#| label: replace-observe-with-observeEvent

observeEvent(r$x, { 
  r$count <- r$count + 1 
}) 
```

To prevent an infinite loop with `observe()` in @lst-15-infinite-loop-observer you could use `isolate()` as in @lst-15-using-isolate, but also  `observeEvent()`
:::

::::
:::::




Indeed, we could used `observeEvent(x, y)` because it is equivalent to` observe({x; isolate(y)})`. It elegantly decouples what you want to listen to from what action you want to take. And `eventReactive()` performs the analogous job for reactives: `eventReactive(x, y)` is equivalent to `reactive({x; isolate(y)})`.

`observeEvent()` and `eventReactive()` have additional arguments that allow you to control the details of their operation:

- By default, both functions will ignore any event that yields NULL (or in the special case of action buttons, 0). Use `ignoreNULL = FALSE` to also handle NULL values.
- By default both functions will run once when you create them. Use `ignoreInit = TRUE` to skip this run.
- For `observeEvent()` only, you can use `once = TRUE` to run the handler only once.


### Exercises {#sec-15-exercises-3}

#### Complete app

Complete the app below with a server function that updates `out` with the value of `x` only when the button is pressed.

:::{.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-complete-app-isolate}
: Complete app using `isolate()`
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-complete-app-isolate}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 400
#| components: [editor, viewer]

## file: app.R
{{< include apps_15/03-ex-complete-app-isolate/app.R >}}
```

App updates with the value of `x` only when the button is pressed
:::

::::
:::::

:::


## Timed invalidation

`isolate()` reduces the time the reactive graph is invalidated. This topic of this section, `invalidateLater()` does the opposite: it lets you invalidate the reactive graph when no data has changed. You saw an example of this in @sec-03-timed-invalidation with `reactiveTimer()`, but the time has come to discuss the underlying tool that powers it: `invalidateLater()`.

`invalidateLater(ms)` causes any reactive consumer to be invalidated in the future, after ms milliseconds. It is useful for creating animations and connecting to data sources outside of Shiny’s reactive framework that may be changing over time. For example, this reactive will automatically generate 10 fresh random normals every half a second:


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-observer-incrementing-number}
: Observer incrementing number
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-observer-incrementing-number}
```{shinylive-r}
#| standalone: true
#| viewerHeight: 300
#| components: [editor, viewer]

library(shiny)

ui <- fluidPage(
  textOutput("sum")
)

server <- function(input, output, session) {

  sum <- reactiveVal(0)
  
  observe({
    invalidateLater(300)
    sum(isolate(sum()) + runif(1))
  })
  
  output$sum <- renderText({
    if (sum() > 10) stop()
    sum()
    })
}

shinyApp(ui, server)
```

Observer increments a cumulative sum with a random number and stops with number over 10 
:::

::::
:::::

In the following sections you’ll learn how to use `invalidateLater()` to read changing data from disk, how to avoid getting `invalidateLater()` stuck in an infinite loop, and some occasionally important details of exactly when the invalidation happens.

### Polling

A useful application of `invalidateLater()` is to connect Shiny to data that is changing outside of R. For example, you could use the following reactive to re-read a csv file every second:


::: {#lst-15-re-reading-csv-file1}

```{{r}}
data <- reactive({
  on.exit(invalidateLater(1000))
  read.csv("data.csv")
})
```

Not run: Re-reading a CSV file with `invalidateLater()`
:::

This connects changing data into Shiny’s reactive graph, but it has a serious downside: when you invalidate the reactive, you’re also invalidating all downstream consumers, so even if the data is the same, all the downstream work has to be redone.

To avoid this problem, Shiny provides `reactivePoll()` which takes two functions: one that performs a relatively cheap check to see if the data has changed and another more expensive function that actually does the computation. We can use `reactivePoll()` to rewrite the previous reactive as below.

::: {#lst-15-re-reading-csv-file3}

```{{r}}
server <- function(input, output, session) {
  data <- reactivePoll(1000, session, 
    function() file.mtime("data.csv"),
    function() read.csv("data.csv")
  )
}
```

Not run: Re-reading a CSV file with `reactivePoll()`
:::

Here we used to `file.mtime()`, which returns the last time the file was modified, as a cheap check to see if we need to reload the file:

Reading a file when it changes is a common task, so Shiny provides an even more specific helper with `reactiveFileReader()` that just needs a file name and a reader function:

::: {#lst-15-re-reading-csv-file3}

```{{r}}
server <- function(input, output, session) {
  data <- reactiveFileReader(1000, session, "data.csv", read.csv)
}
}
```

Not run: Re-reading a CSV file with `reactiveFileReader()`
:::

If you need to read changing data from other sources (e.g. a database), you’ll need to come up with your own `reactivePoll()` code.


### Long running reactives

If you’re performing a long running computation, there’s an important question you need to consider: when should you execute `invalidateLater()`? For example, take this reactive:

::: {#lst-15-long-running-reactive1}

```{{r}}
server <- function(input, output, session) {
  data <- reactiveFileReader(1000, session, "data.csv", read.csv)
}
```

Not run: Long running reactive with `invalidateLater()` time problem.
:::

Assume Shiny starts the reactive running at time 0, it will request invalidation at time 500. The reactive takes 1000ms to run, so it’s now time 1000, and it’s immediately invalidated and must be recomputed, which then sets up another invalidation: we’re stuck in an infinite loop.

On the other hand, if you run `invalidateLater()` at the end, it will invalidate 500ms after completion, so the reactive will be re-run every 1500 ms.

::: {#lst-15-long-running-reactive2}

```{{r}}
x <- reactive({
  on.exit(invalidateLater(500), add = TRUE)
  Sys.sleep(1)
  10
})
```

Not run: Long running reactive with `invalidateLater()` time problem solved.
:::

This is the main reason to prefer `invalidateLater()` to the simpler `reactiveTimer()` that we used earlier: it gives you greater control over exactly when the invalidation occurs.

### Timer accuracy

The number of milliseconds specified in `invalidateLater()` is a polite request, not a demand. R may be doing other things when you asked for invalidation to occur, so your request has to wait. This effectively means that the number is a minimum and invalidation might take longer than you expect. In most cases, this doesn’t matter because small differences are unlikely to affect user perception of your app. However, in situations where many small errors will accumulate, you should compute the exact elapsed time and use it to adjust your calculations.

For example, the following code computes distance based on velocity and elapsed time. Rather than assuming `invalidateLater(100)` always delays by exactly 100 ms, we’ll compute the elapsed time and use it in the calculation of position.


::: {#lst-15-timer-accuracy}

```{{r}}
velocity <- 3
r <- reactiveValues(distance = 1)

last <- proc.time()[[3]]
observe({
  cur <- proc.time()[[3]]
  time <- last - cur
  last <<- cur
  
  r$distance <- isolate(r$distance) + velocity * time
  invalidateLater(100)
})
```


Not run: Computing elapsed time for timer accuracy
:::




### Exercises {#sec-15-exercises4}

#### Why reactive never executed?

Why will this reactive never be executed? Your explanation should talk about the reactive graph and invalidation.


:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-ex-reactive-never-executed1}
: Exercise: A reactive never executed
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-ex-reactive-never-executed1}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_15/04-ex-reactive-never-executed1/app.R >}}
```

Exercise: The reactive is never executed because there is no output/observer and input/value to call the reactive expression invalidate. Shiny can't therefore start (re)-execution.
:::

::::
:::::


The reactive is never called. There is no input that can be invalidated and on the other hand, there is no reactive consumer (an output) to display changed values. 

After initializing a session, Shiny can't pick an observer because there is none. As Shiny does not choose a reactive expression for the execution start, and there is no input that could be invalidated, the reactive can't never be called. 

To see the difference I have complemented the app in @lst-15-ex-reactive-never-executed1 with an output. Now you can see that the reactive is executed.

:::{.column-page-inset}

:::::{.my-r-code}
:::{.my-r-code-header}
:::::: {#cnj-15-ex-reactive-never-executed2}
: Exercise: The reactive is this time executed
::::::
:::
::::{.my-r-code-container}
::: {#lst-15-ex-reactive-never-executed2}

```{shinylive-r}
#| standalone: true
#| viewerHeight: 550
#| components: [editor, viewer]

## file: app.R
{{< include apps_15/04-ex-reactive-never-executed2/app.R >}}
```

Exercise: The reactive can now be executed, because Shiny can invalidate the output/observer and the reactive can be called.
:::

::::
:::::

:::

::: {.callout-caution #cau-stop-app-not-working-with-shinylive-r}
###### `stopApp()` not working with `shinylive-r`

I had to use a more complex logic with an intermediate reactive value because the simpler solution with `stopApp()` inside an `observeEvent()` listening to the "stop" `actionButton()` does not work with `shinylive-r`.
:::

#### `reactivePoll()` with SQL

If you’re familiar with SQL, use `reactivePoll()` to only re-read an imaginary “Results” table whenever a new row is added. You can assume the Results table has a `timestamp` field that contains the date-time that a record was added.

As I am not familiar with `r glossary("SQL")` I will skip this exercise.


## Glossary Entries {#unnumbered}

```{r}
#| label: glossary-table
#| echo: false

glossary_table()
```

------------------------------------------------------------------------

## Session Info {.unnumbered}

::: my-r-code
::: my-r-code-header
Session Info
:::

::: my-r-code-container
```{r}
#| label: session-info

sessioninfo::session_info()
```
:::
:::

